---
title: "bayesian"
editor: visual
---

```{r, include=FALSE}
library(tidyverse)
library(dplyr)
library(rempsyc)
library(psych)
library(ROCR)
library(janitor)
library(tidybayes)
library(ggplot2)
library(rstanarm)
library(bayesrules)
library(bayesplot)
library(loo)
```

##  Raw data

### Data Origin

The dataset is obtained from UCI Repository, collected at 'Hospital Universitario de Caracas' in Caracas, Venezuela. The dataset comprises demographic information, habits, and historic medical records of 858 patients.

```{r, include=FALSE, warning=FALSE}
url <- "https://raw.githubusercontent.com/mwc201/BayesianCancerPrediction/main/data/kag_risk_factors_cervical_cancer.csv"

cancer_full <- read.csv(url)

# change all ? to NA

cancer_full <- cancer_full %>%

  mutate(across(everything(), sub, pattern = "?", replacement = NA, fixed = TRUE))

# Convert data type for predictor variables

cancer_full <- cancer_full %>%

  mutate(Age = as.integer(Age),

         Number.of.sexual.partners = as.integer(Number.of.sexual.partners),

         Hormonal.Contraceptives = as.factor(as.integer(Hormonal.Contraceptives)),

         IUD = as.factor(as.integer(IUD)),

         Smokes = as.factor(as.integer(Smokes)),

         STDs = as.factor(as.integer(STDs)),

         Dx = as.factor(Dx),

         Hinselmann = as.factor(Hinselmann),

         Schiller = as.factor(Schiller),

         Citology = as.factor(Citology),

         Biopsy = as.factor(Biopsy))

# Select the desired columns

cancer <- cancer_full %>%

  select(c(Age, Number.of.sexual.partners, Hormonal.Contraceptives,

           IUD, Smokes, STDs, Hinselmann, Schiller, Citology,

           Biopsy))

cancer <- na.omit(cancer)
```

###  Target Variable Selection

Hinselmanns test refers to colposcopy using acetic acid. Meanwhile, colposcopy using Lugoliodine includes Schillers test, Cytology and Biopsy. Positive exams results don't necessarily imply a diagnostic, but as multiple exams return positive, the greater the likelihood of cervical cancer).

In this project, we selected only one target variable to measure cervical cancer. We conducted separate logistic regressions on all of the original target variables (Hinselmann, Schiller, Citology, and Biopsy) and see which one would be the best for cervical cancer measurement.

```{r, include=TRUE,warning=FALSE}
# use auc scores to test which model predicts the outcomes the best  

set.seed(1111)

auc1 <- vector()

auc2 <- vector()

auc3 <- vector()

auc4 <- vector()

train <- data.frame()

test <- data.frame()

selected_rows <- replicate(100,sample(1:nrow(cancer),ceiling(nrow(cancer)*0.5),

                            replace=FALSE))

selected_rows <- as.data.frame(selected_rows)

for (i in 1:100) {

  train <- cancer[selected_rows[,i], ]

  test <- cancer[-selected_rows[,i], ]

  

  fit1 <- glm(Hinselmann ~ Age + Number.of.sexual.partners + Hormonal.Contraceptives +

              IUD + Smokes + STDs, data = train, family = "binomial")

  test$predicted.prob1 <- predict(fit1, test, type = "response")

  pred1 <- prediction(test$predicted.prob1, test$Hinselmann)

  perf1 <- performance(pred1, "auc")

  auc1[i] <- round(perf1@y.values[[1]], 3)

  

  fit2 <- glm(Schiller ~ Age + Number.of.sexual.partners + Hormonal.Contraceptives +

              IUD + Smokes + STDs, data = train, family = "binomial")

  test$predicted.prob2 <- predict(fit2, test, type = "response")

  pred2 <- prediction(test$predicted.prob2, test$Schiller)

  perf2 <- performance(pred2, "auc")

  auc2[i] <- round(perf2@y.values[[1]], 3)

  

  fit3 <- glm(Citology ~ Age + Number.of.sexual.partners + Hormonal.Contraceptives +

              IUD + Smokes + STDs, data = train, family = "binomial")

  test$predicted.prob3 <- predict(fit3, test, type = "response")

  pred3 <- prediction(test$predicted.prob3, test$Citology)

  perf3 <- performance(pred3, "auc")

  auc3[i] <- round(perf3@y.values[[1]], 3)

  fit4 <- glm(Biopsy ~ Age + Number.of.sexual.partners + Hormonal.Contraceptives +

              IUD + Smokes + STDs, data = train, family = "binomial")

  test$predicted.prob4 <- predict(fit4, test, type = "response")

  pred4 <- prediction(test$predicted.prob4, test$Biopsy)

  perf4 <- performance(pred4, "auc")

  auc4[i] <- round(perf4@y.values[[1]], 3)

}

df <- data.frame(

  Target_Variable = c("Hinselmann", "Schiller", "Citology", "Biopsy"),

  AUC = c(mean(auc1), mean(auc2), mean(auc3), mean(auc4))

)
```

```{r, include=FALSE}
# Table 1: AUC 

knitr::kable(df)
```

As Table 1 shown, Schiller has the highest AUC value. In this case, we selected Schiller as the target variable. Following this, we would conduct an EDA on risk factors for cervical cancer leading to a Schiller Examination.

```{r}
# drop other target variables

cancer <- cancer %>% select(-Hinselmann, -Citology, -Biopsy)

```

### Exploratory Data Analysis

\*\*Schiller\*\* Taking values between 0 (negative) or 1 (positive). 644 out of 712 observations (90.5%) were negative for abnormal cells, while the remaining 68 observations (9.5%) were positive.

\*\*Age\*\* The age of respondents. Values ranged from 13 to 84. We computed descriptive statistics (M = 27.25; SD = 8.77; skewness = 1.40; kurtosis = 4.73).

```{r fig.dim=c(5,3), out.width="70%", fig.align = 'center'}
cancer %>% ggplot(aes(x = Age, fill = factor(Schiller))) +

  geom_histogram(alpha = 0.5, position = "identity", bins = 20) +

  scale_fill_manual(values = c("#619CFF", "#FF595E"), name = "Schiller") +

  labs(x = "Age", y = "Count") +

  ggtitle("Histogram of Age by Schiller") +

  theme_minimal()
```

\*\*Number.of.sexual.partners\*\* Number of sexual partners. Values ranged from 1 to 28. We computed descriptive statistics (M = 2.51; SD = 1.64; skewness = 5.90; kurtosis = 81.68). On average, we could conclude that patients have had 2-3 sexual partners.

```{r  fig.dim=c(5,3), out.width="70%", fig.align = 'center',warning=FALSE}
cancer %>% ggplot(aes(x = Number.of.sexual.partners, fill = factor(Schiller))) +

  geom_histogram(alpha = 0.5, position = "identity", bins = 20) +

  scale_fill_manual(values = c("#619CFF", "#FF595E"), name = "Schiller") +

  scale_x_continuous(limits = c(0, 10), breaks = seq(0, 10, 1)) +

  labs(x = "Number of Sexual Partners", y = "Count") +

  ggtitle("Histogram of Number of Sexual Partners by Schiller") +

  theme_minimal()

```

```{r categorical_v, include=FALSE}
#summary statistics for the categorical variables

summary(cancer[c("Schiller", "Hormonal.Contraceptives", "IUD", "Smokes", "STDs")])
```

\*\*Hormonal.Contraceptives\*\* Whether use hormonal contraceptives or not. Taking values between 0 (negative) or 1 (positive). 254 out of 712 observations (35.7%) were negative for abnormal cells, while the remaining 458 observations (64.3%) were positive.

\*\*IUD\*\* Whether use the intrauterine device (hormonal control method) or not. Taking values between 0 (negative) or 1 (positive). 631 out of 712 observations (88.6%) not use intrauterine device, while the remaining 68 observations (11.4%) use intrauterine device.

\*\*Smokes\*\* Whether smokes or not. Taking values between 0 (non-smoke) or 1 (smoke). 609 out of 712 observations (85.5%) were non-smokers, while the remaining 103 observations (14.5%) were smokers.

\*\*STDs\*\* Whether have sexually transmitted diseases or not. Taking values between 0 (negative) or 1 (positive). 645 out of 712 observations (90.6%) were negative for abnormal cells, while the remaining 67 observations (9.4%) were positive.

```{r plot_c, fig.dim=c(5,3), out.width="80%", fig.align = 'center'}
#distribution of the categorical variables

cancer %>% select(Hormonal.Contraceptives, IUD, Smokes, STDs) %>% 

  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%

  ggplot(aes(x = factor(value))) +

  geom_bar() +

  facet_wrap(~ variable, scales = "free_x", ncol = 2) +

  labs(title = "Distribution of Categorical Variables") +

  xlab("") +

  ylab("Count") +

  theme_minimal()
```

We can observe that:

\- About 64% of the patients have used hormonal contraceptives.

\- About 11% of the patients have used IUD.

\- About 15% of the patients smoke.

\- About 9.4% of the patients have a history of sexually transmitted diseases.

\*\*Correlation\*\* We can examine the correlation among the key study variables according to following Table 2.

```{r corr, include=FALSE}
corstarsl <- function(x){ 

  require(Hmisc) 

  x <- as.matrix(x) 

  R <- rcorr(x)$r 

  p <- rcorr(x)$P

  # define notions for significance levels; spacing is important.

  mystars <- ifelse(p < .001, "***", ifelse(p < .01, "** ", ifelse(p < .05, "* ", " ")))

  # trunctuate the matrix that holds the correlations to two decimal

  R <- format(round(cbind(rep(-1.11, ncol(x)), R), 2))[,-1] 

  # build a new matrix that includes the correlations with their apropriate stars 

  Rnew <- matrix(paste(R, mystars, sep=""), ncol=ncol(x)) 

  diag(Rnew) <- paste(diag(R), " ", sep="") 

  rownames(Rnew) <- colnames(x) 

  colnames(Rnew) <- paste(colnames(x), "", sep="") 

  # remove upper triangle

  Rnew <- as.matrix(Rnew)

  Rnew[upper.tri(Rnew, diag = TRUE)] <- ""

  Rnew <- as.data.frame(Rnew) 

  # remove last column and return the matrix (which is now a data frame)

  Rnew <- cbind(Rnew[1:length(Rnew)-1])

  return(Rnew) 

}
```

```{r corr_tbl, message=FALSE}
# Table 2: correlation matrix

cancer_table2 <- cancer %>% rename("Num.Sex.Par"=Number.of.sexual.partners) %>% 

  rename("HC"=Hormonal.Contraceptives)

knitr::kable(corstarsl(cancer_table2))
```

All the correlation values are very low , so our model would not have the problem of multicollinearity.

```{r plot, fig.dim=c(5,3), out.width="80%", warning=FALSE, fig.align = 'center'}
# removing an outlier (28)

cancer %>% ggplot(aes(Age, Number.of.sexual.partners)) +

  geom_point() +

  scale_y_continuous(limits = c(0, 10)) +

  ylab("Number of Sexual Partners") +

  theme_minimal()
```

We can observe that the number of sexual partners reach the peak in the 20-30 age group and then gradually decreases with age.

## Prior Predictive Simulation

To complete the Bayesian logistic regression model of Y, we must put prior models on our regression parameters. As usual, since these parameters can take any value in the real line, Normal priors are appropriate for both. We'll also assume independence among the priors and express our prior understanding of the model baseline beta_o through the centered intercept beta_0c:

According to the research, the expected log(odds) of the coefficient ranges from -0.1 to 1. Thus we set our prior to a normal distribution with mean 0.5 and a standard error of 0.3.

Next, we will use informative priors for the coefficients of other variables in the dataset. Then we simulated data under a variety of prior models.

```{r, results='hide', include=FALSE}
cancer_model_prior <- stan_glm(Schiller~Age

                               +Number.of.sexual.partners

                               +Hormonal.Contraceptives

                               +IUD+Smokes+STDs,

                               data = cancer, family = binomial,

                               prior_intercept = normal(-3, 0.3),

                               prior = normal(0.5, 0.3),

                               chains = 4, iter = 5000*2, seed = 1,

                               prior_PD = TRUE)
```

```{r, warning=FALSE, message=FALSE, fig.show="hold", out.width="50%", fig.align = 'center'}
set.seed(11)

# plot 100 prior models, look at the prediction on the probability scale

cancer_epred <- cancer %>% 

  add_epred_draws(cancer_model_prior, ndraws = 100) 

cancer_epred %>%

  ggplot(aes(y = Schiller, x = Smokes)) +

  geom_line(aes(y = .epred, group = .draw), size = 0.1) +

  theme_minimal()

cancer_epred %>%

  ggplot(aes(y = Schiller, x = STDs)) +

  geom_line(aes(y = .epred, group = .draw), size = 0.1) +

  theme_minimal()
```

```{r, warning=FALSE, message=FALSE, fig.show="hold", out.width="50%", fig.align = 'center'}
# plot the observed proportion of positive results in 100 prior datasets

cancer %>% 

  add_predicted_draws(cancer_model_prior, n = 100) %>% 

  group_by(.draw) %>% 

  summarise(proportion_cancer = mean(.prediction == 1)) %>% 

  ggplot(aes(x = proportion_cancer)) +

  geom_histogram(color = "white") +

  theme_minimal()
```

We plot 100 of these prior plausible relationships as shown above. These adequately reflect our prior understanding that the individuals who smoke or have sexually transmitted diseases have higher probability of getting positive Schiller results than the individuals who do not smoke or have diseases, as well as our prior uncertainty around the rate of this increase. The histogram of the 100 predicted positive result proportions from our 100 prior simulated datasets is displayed. We can observe that the prior predictions tend to be centered around low values. It indicates that our prior tuning is reasonable since we believe that the ratio of getting positive Schiller results should be low. Furthermore, the percent of getting positive results ranged from as low as roughly 0.03 in one dataset to as high as roughly 0.37 in another. This does adequately match our prior understanding and uncertainty about getting positive Schiller results.
